---
title: "Forecasting Red Snapper"
author: "Andrew Rafael James"
date: "2025-03-02"
output: html_document
---

# Load Dataset
```{r setup, warning = FALSE, message = FALSE}
library(dplyr)
library(fpp)
library(forecast)
library(tidyverse)

# Load red_snapper data
load("/Users/andrewrafaeljames/Documents/Time Series and Forecasting/Group Project/Data/red_snapper.rda")

head(red_snapper)
```


# Transform red_snapper into time-series object

summary(rs_ts)

``` {r transform}
library(tsibble)
rs_ts = ts(red_snapper, frequency = 12, start = c(2007, 1))

# Plot Monthly Number of Red Snapper Harvested in Gulf of Mexico
# Data shows increasing trend with seasonality. Variance also looks increasing overtime
autoplot(rs_ts[, 'landings']) + labs(title = 'Monthly Number of Red Snapper Landings\n in Gulf of Mexico between 2007-2023',
                                     y = '# of Landings') + theme(plot.title = element_text(size = 14, hjust = 0.5),
        plot.margin = margin(t = 15, r = 10, b = 10, l = 10))

test = ts(red_snapper, frequency = 12, start = c(2007, 1), end = c(2009, 12))
autoplot(test[, 'landings']) + labs(title = 'Monthly Number of Red Snapper Landings in Gulf of Mexico between 2007-2023',
                                     y = '# of Landings')

ggseasonplot(rs_ts[, 'landings'], continuous = TRUE, main = 'Seasonal Plot of Number of Red Snapper Landings Across Months') + labs(y = 'landings') + theme(plot.title = element_text(size = 14, hjust = 0.5),
        plot.margin = margin(t = 15, r = 10, b = 10, l = 10))
ggsubseriesplot(rs_ts[, 'landings'], main = 'Seasonal Subseries Plot of Number of Red Snapper Landings') + labs(y = 'landings') + theme(plot.title = element_text(size = 14, hjust = 0.5),
        plot.margin = margin(t = 15, r = 10, b = 10, l = 10))
```

# Moving Average Smoothing

``` {r moving_average_smoothing}
library(TTR)
library(gridExtra)

sma_4 = SMA(rs_ts[, 'landings'], n = 4)
sma_6 = SMA(rs_ts[, 'landings'], n = 6)
sma_8 = SMA(rs_ts[, 'landings'], n = 8)
sma_12 = SMA(rs_ts[, 'landings'], n = 12)
sma_16 = SMA(rs_ts[, 'landings'], n = 16)


time_index = time(rs_ts)
ma = data.frame(time = time_index, landings = rs_ts[, 'landings'], sma_4 = as.numeric(sma_4),
               sma_6 = as.numeric(sma_6), sma_8 = as.numeric(sma_8), sma_12 = as.numeric(sma_12))
p1 = ggplot(data = ma, aes(x = time)) + geom_line(aes(y = landings), color = 'grey') +
  geom_line(aes(y = sma_4), color = 'red') + labs(title = "4-Month Moving Average Smoothing") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5, margin = margin(b = 10)))

p2 = ggplot(data = ma, aes(x = time)) + geom_line(aes(y = landings), color = 'grey') +
  geom_line(aes(y = sma_6), color = 'red') + labs(title = "6-Month Moving Average Smoothing") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5, margin = margin(b = 10)))

p3 = ggplot(data = ma, aes(x = time)) + geom_line(aes(y = landings), color = 'grey') +
  geom_line(aes(y = sma_8), color = 'red') + labs(title = "8-Month Moving Average Smoothing") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5, margin = margin(b = 10)))

p4 = ggplot(data = ma, aes(x = time)) + geom_line(aes(y = landings), color = 'grey') +
  geom_line(aes(y = sma_12), color = 'red') + labs(title = "12-Month Moving Average Smoothing") +
  theme(plot.title = element_text(size = 11, face = "bold", hjust = 0.5, margin = margin(b = 10)))

grid.arrange(p1, p2, p3, p4, ncol = 2)



```
# Rolling Standard Deviation
``` {r rolling_std}
rolling_std = runSD(rs_ts[, 'landings'], n = 12)

# Plot original data and rolling standard deviation
plot(rolling_std, main = "Rolling Standard Deviation", ylab = 'Standard Deviation', xlab = 'Time', lwd = 2, col = 'blue')
```


# Plot Monthly Allocation and Vessel Price in Gulf of Mexico
``` {r plot_xvars}
autoplot(rs_ts[, c('allocation_price', 'vessel_price')]) + labs(title = 'Monthly Allocation and Vessel Price\n in Gulf of Mexico between 2007-2023', y = 'Price') + theme(plot.title = element_text(size = 14, hjust = 0.5),
        plot.margin = margin(t = 15, r = 10, b = 10, l = 10))
```


# Show time-series data along with ACF and PACF
Interpretation ACF shows significant lags decaying slowly, indicating non stationarity
``` {r acf_pacf}
# Filter landing column only
landing_ts = rs_ts[, 'landings']
tsdisplay(landing_ts, main = 'ACF and PACF Plot of # of Landings')
```

# Decompose TS into trend, seasonality
Interpretation: Since variance looks increasing overtime, Holt-Winter Multiplicative Method should be considered

``` {r decomposition}
decomposed_ts = decompose(landing_ts, type = 'multiplicative')

# library(ggseas)
library(ggplot2)

autoplot(decomposed_ts) + 
  ggtitle("Decomposing Time-Series into \nSeasonal, Trend, and Remainder Components") +
  theme(plot.title = element_text(size = 14, hjust = 0.5),
        plot.margin = margin(t = 15, r = 10, b = 10, l = 10))


# plot(decompose(landing_ts, type = 'additive'))
```

# Split data into train test split
``` {r split_data}
train_df = window(landing_ts, frequency = 12, start = c(2007, 1), end = c(2022, 12))
test_df = window(landing_ts, frequency = 12, start = c(2023, 1))
```

# EXPONENTIAL SMOOTHING METHOD

``` {r check_lambda}
# lambda = BoxCox.lambda(train_df)
lambda = BoxCox.lambda(train_df)
train_df_boxcox = BoxCox(train_df, lambda = lambda)
```


# Holts-Winter Additive
``` {r hw_additive}
hw_additive = hw(train_df_boxcox, h = 12, seasonal = "additive", level = c(80, 95))
forecast_hw_additive = hw_additive$mean
summary(hw_additive)
```

# Holts-Winter Multiplicative
``` {r hw_multiplicative}
hw_multiplicative = hw(train_df, h = 12, seasonal = "multiplicative", level = c(80, 95))
forecast_hw_multiplicative = hw_multiplicative$mean
summary(hw_multiplicative)

```

# Holts-Winter Additive with Damping
``` {r hw_additive_damping}
hw_additive_damping = hw(train_df_boxcox, h = 12, seasonal = "additive", damped = TRUE, level = c(80, 95))
forecast_hw_additive_damping = hw_additive_damping$mean
summary(hw_additive_damping)

```

# Holts-Winter Multiplicative with Damping
``` {r hw_multiplicative_damping}
hw_multiplicative_damping = hw(train_df, h = 12, seasonal = "multiplicative", damped = TRUE, level = c(80, 95))
forecast_hw_multiplicative_damping = hw_multiplicative_damping$mean
summary(hw_multiplicative_damping)
```

# Exponential Trend Holts-Winter Multiplicative with Damping
``` {r hw_exp_multiplicative_damping}
hw_exp_multiplicative_damping = hw(train_df, h = 12, seasonal = "multiplicative", damped = TRUE, exponential = TRUE, level = c(80, 95))
forecast_hw_exp_multiplicative_damping = hw_exp_multiplicative_damping$mean
summary(hw_exp_multiplicative_damping)
```
# ETS
``` {r ets}
ets_model = ets(train_df)
forecast_ets_model = forecast(ets_model, h = 12)$mean
summary(ets_model)
```

# Seasonal Naive Method
``` {r snaive}
snaive = snaive(train_df_boxcox, h = 12)
forecast_snaive = snaive$mean
summary(snaive)
```

# Check accuracy agains test set

summary(hw_exp_multiplicative_damping)$aic

``` {r check_accuracy}
options(width = 120)

accuracy_hw_additive = accuracy(InvBoxCox(forecast_hw_additive, lambda), test_df)
accuracy_hw_multiplicative = accuracy(forecast_hw_multiplicative, test_df)
accuracy_hw_additive_damping = accuracy(InvBoxCox(forecast_hw_additive_damping, lambda), test_df)
accuracy_hw_multiplicative_damping = accuracy(forecast_hw_multiplicative_damping, test_df)
accuracy_hw_exp_multiplicative_damping = accuracy(forecast_hw_exp_multiplicative_damping, test_df)
accuracy_ets = accuracy(forecast_ets_model, test_df)
accuracy_snaive = accuracy(InvBoxCox(forecast_snaive, lambda), test_df)

result = as.data.frame(rbind(accuracy_hw_additive, 
                    accuracy_hw_multiplicative, 
                    accuracy_hw_additive_damping, 
                    accuracy_hw_multiplicative_damping,
                    accuracy_hw_exp_multiplicative_damping, 
                    # accuracy_ets,
                    accuracy_snaive))

rownames(result) = c('HW Additive', 'HW Multiplicative', 'HW Additive with Damping',
                     'HW Multiplicative with Damping', 'Exponential HW Multiplicative with Damping',
                     # 'ETS (MAM)',
                     'Seasonal Naive')


result$AICc = c(2744.608, 5394.614, 2747.643, 5384.184, 5381.935, 'N/A')
result %>% select(AICc, everything())

plot(hw_additive_damping)
```

# Model Diagnostic
``` {r check_model_diagnostic}
checkresiduals(hw_additive_damping)
```


# ARIMA METHOD
``` {r arima}
tsdisplay(train_df)
lambda = BoxCox.lambda(train_df)
train_df_boxcox = BoxCox(train_df, lambda = lambda)

train_df_boxcox
autoplot(train_df_boxcox) + labs(title = 'Monthly Number of Red Snapper Harvested in Gulf of Mexico between 2007-2023',
                                     y = '# of Landings')

train_df_boxcox_diff = diff(train_df_boxcox, lag = 6)
tsdisplay(train_df_boxcox_diff)

kpss.test(train_df_boxcox_diff)

auto_arima_model = auto.arima(train_df, lambda = lambda, d = 1, D = 1, seasonal = TRUE)
auto_arima_model_2 = auto.arima(train_df, lambda = lambda, seasonal = TRUE)

auto_arima_model
auto_arima_model_2

checkresiduals(auto_arima_model)

auto_arima_model_forecast = forecast(auto_arima_model, h = 12)
forecast_sarima = auto_arima_model_forecast$mean

accuracy(forecast_sarima, test_df)
```

# TIME-SERIES REGRESSION

hist(train_df, main="Histogram of Landings", col="blue")
hist(train_df_all[, 'allocation_price'], main="Histogram of Landings", col="blue")
hist(train_df_all[, 'vessel_price'], main="Histogram of Landings", col="blue")

``` {r timeseries_regression}
options(width = 120)
log_rs_ts = log(rs_ts)
cor(rs_ts)
cor(log_rs_ts)

train_df_all = window(log_rs_ts, frequency = 12, start = c(2007, 1), end = c(2022, 12))
test_df_all = window(log_rs_ts, frequency = 12, start = c(2023, 1))

train_df_all

autoplot(train_df_all[, c('landings', 'allocation_price', 'vessel_price')]) + labs(title = 'Monthly Number of Red Snapper Harvested in Gulf of Mexico between 2007-2023',
                                     y = '# of Landings')


autoplot(train_df_all[, 'allocation_price']) + labs(title = 'Monthly Allocation Price in Gulf of Mexico between 2007-2023', 
                                                    y = 'Allocation Price')

autoplot(train_df_all[, 'vessel_price']) + labs(title = 'Monthly Vessel Price in Gulf of Mexico between 2007-2023', 
                                                    y = 'Vessel Price')

linear_model = tslm(train_df_all[, 'landings'] ~ train_df_all[, 'allocation_price'] + train_df_all[, 'vessel_price'])
linear_model

combined = data.frame(landings = as.numeric(train_df_all[, 'landings']), 
           allocation_price = as.numeric(train_df_all[, 'allocation_price']), 
           vessel_price = as.numeric(train_df_all[, 'vessel_price']))

combined_df = combined %>% pivot_longer(names_to = 'variable', values_to = 'value', cols = c('allocation_price', 'vessel_price'))

combined_df

ggplot(data = combined, aes(x = landings, y = allocation_price)) + geom_point() + geom_smooth(method = 'lm', color = 'red') + labs(title = 'Landings vs Allocation Price with Fitted Trend)', x = 'Landings', y = 'Allocation Price')

ggplot(data = combined_df, aes(x = landings, y = value, color = variable)) + geom_point() + geom_smooth(method = 'lm', color = 'red')

summary(linear_model)
checkresiduals(linear_model)
```



# REGRESSION WITH ARMA ERROR


``` {r regression_with_arma_error}
armax_all = auto.arima(train_df_all[, 'landings'], xreg = train_df_all[, 'allocation_price'], D = 1)
armax_all

armax_ves = auto.arima(train_df_all[, 'landings'], xreg = train_df_all[, 'vessel_price'], D = 1)
armax_ves

armax_all_ves = auto.arima(train_df_all[, 'landings'], xreg = train_df_all[, c('allocation_price','vessel_price')], D = 1)
armax_all_ves

checkresiduals(armax_all)

checkresiduals(armax_ves)
checkresiduals(armax_all_ves)

# forecast of exogeneous variables
naive_vessel_forecast = naive(train_df_all[, 'vessel_price'], h = 12)

# Vessel price
vessel_arima = auto.arima(train_df_all[, 'vessel_price'], seasonal = TRUE)
forecast_vessel_arima = forecast(vessel_arima, h = 12)

snaive_vessel = snaive(train_df_all[, 'vessel_price'])
forecast_vessel_snaive = forecast(snaive_vessel, h = 12)

accuracy(forecast_vessel_arima$mean, test_df_all[, 'vessel_price'])
accuracy(forecast_vessel_snaive$mean, test_df_all[, 'vessel_price'])

# Allocation Price
allocation_arima = auto.arima(train_df_all[, 'allocation_price'], seasonal = TRUE)
forecast_allocation_arima = forecast(allocation_arima, h = 12)

snaive_allocation = snaive(train_df_all[, 'allocation_price'])
forecast_allocation_snaive = forecast(snaive_allocation, h = 12)

accuracy(forecast_allocation_arima$mean, test_df_all[, 'allocation_price'])
accuracy(forecast_allocation_snaive$mean, test_df_all[, 'allocation_price'])

# Landings and Vessel Price
armax_ves_forecast = forecast(armax_ves, xreg = forecast_vessel_arima$mean, h = 12)
armax_ves_forecast$mean
accuracy(exp(armax_ves_forecast$mean), test_df)
# accuracy(armax_ves_forecast$mean, test_df)


# Landings, Allocation Price
armax_all_forecast = forecast(armax_all, xreg = forecast_allocation_snaive$mean, h = 12)
armax_all_forecast
accuracy(exp(armax_all_forecast$mean), test_df)
# accuracy(armax_all_forecast$mean, test_df)


# Landings, Allocation Price, and Vessel Price
armax_all_vessel_forecast = forecast(armax_all_ves, xreg = cbind(forecast_allocation_snaive$mean, forecast_vessel_arima$mean), h = 12)
armax_all_vessel_forecast
accuracy(exp(armax_all_vessel_forecast$mean), test_df)
# accuracy(armax_all_vessel_forecast$mean, test_df)

```

# VAR
``` {r var}
library(vars)

train_df_all

var_model = VAR(train_df_all[, c('landings', 'allocation_price', 'vessel_price')], type = 'both', season = 12, p = 3 )
summary(var_model)

acf(residuals(var_model))

var_forecast = predict(var_model, n.ahead = 12)
var_forecast$fcst$landings[, 1]

accuracy(var_forecast$fcst$landings[, 1], test_df)
```

# SPECTRAL
``` {r spectral}
library(TSA)
periodogram = periodogram(train_df, main = 'Spectral Analysis')

periodogram$freq

Tperiod = (1 / periodogram$freq) / 12

Tperiod
192/12

96/12

64/12
```

